Modeling the Linux page cache for accurate simulation of data-intensive applications

### 2020-12-16 paper_io_sim.tex
integrated real application results in text

        Our experiments compare real executions with our Python prototype,
        with the original \wrench, and with our \wrench-cache extension. We
        evaluate our page cache simulation in single-threaded and
        multi-threaded applications, accessing data on local and remote
        file systems. We use two applications: a synthetic one, designed to
        evaluate the simulation model, and a real one, representative of
        neuroimaging data processing.

        The synthetic application, implemented in C, consists of three single-core,
        sequential tasks where each task reads the file produced by the
        previous task, increments every byte of this file to emulate real
        processing, and writes the resulting data to disk. Files are
        numbered by ascending creation times: File 1 is the file read by Task 1, and file \texttt{i+1} is the file
        written by task \texttt{i}. In real executions, we clear the page
        cache before starting the application, to ensure comparable
        conditions among the runs. The anonymous memory used by application
        is released after each task, which we also simulate in the Python
        prototype and in WRENCH-cache.

        The real application is a workflow of the Nighres
        toolbox~\cite{huntenburg2018nighres} that implements
        cortical reconstruction from brain images in four steps: skull stripping, tissue classification, 
        region extraction, and cortical reconstruction. More information 
        on this applicatoin is available in the Nighres documentation at \url{https://nighres.readthedocs.io/en/latest/auto_examples/example_02_cortical_depth_estimation.html}.
        The workflow is implemented as a Python
        script that calls Java image-processing routines. We used Python
        3.6, Java 8, and Nighres 1.3.0. We patched the
        application to remove lazy data loading, which made CPU time difficult to separate from I/O time, 
        and to monitor task CPU times, to inject them in the simulators. The patched code is 
        available at \url{https://github.com/dohoangdzung/nighres}.

        The applications were executed on a dedicated cluster at
        Concordia University, with one login node, 9 compute nodes, and 4
        storage nodes connected with a network bandwidth of 25 Gbps. Each
        compute node has 2 $\times$ 16-core Intel(R) Xeon(R) Gold 6130 CPU
        @ 2.10GHz, 250~GiB of RAM, 6 $\times$ SSDs of 450~GiB each with the XFS
        file system, 378~GiB of tmpfs, and 126~GiB of devtmpfs file system.
        Nodes run CentOS~8.1 with NFS version 4. We used the \texttt{atop}
        and \texttt{collectl} tools to monitor and collect memory status
        and disk throughput.

        To parameterize the simulators, we benchmarked the
        memory, local disk, remote disk (NFS), and network bandwidths
        (Table~\ref{table:benchmark}). Since \simgrid, and thus \wrench, currently only supports
        symmetrical bandwidths, we use the mean of the read and write
        bandwidth values in our experiments.
            \begin{table}
            \centering
            \begin{tabularx}{\columnwidth}{ll
            >{\centering\arraybackslash}X
            >{\centering\arraybackslash}X
            >{\centering\arraybackslash}X}
            \toprule
                \multicolumn{2}{c}{Bandwidths}  & Cluster (real) & Python prototype & \wrench simulator\\
            \midrule
            \multirow{2}{*}{Memory}      & read  & 6860 & 4812 & 4812\\
                                         & write & 2764 & 4812 & 4812\\
            \multirow{2}{*}{Local disk}  & read  & 510  & 465  & 465\\
                                         & write & 420  & 465  & 465\\
            \multirow{2}{*}{Remote disk} & read  & 515  & -    & 445\\
                                         & write & 375  & -    & 445\\
            \multicolumn{2}{l}{Network}  & 3000  & -    & 3000\\
            \bottomrule
            \end{tabularx}
            \caption{Bandwidth benchmarks (MBps) and simulator configurations.
            The bandwidths used in the simulations are the average of the read and write bandwidths
            measured in the cluster.
            Remote disk and network accesses are not simulated in the Python simulator.}
            \label{table:benchmark}
            \end{table}

            As our focus is on I/O rather than compute, we measured
            application task CPU times on a cluster node
            (Table~\ref{table:cputime}), and used these durations in our
            simulations. For the Python prototype, we injected CPU times
            directly in the simulation. For \wrench and \wrench-cache, we
            determined the corresponding number of flops on a 1~Gflops CPU
            and used these values in the simulation. The final simulated
            platform and application used in our experiments are available at
            commit \href{https://github.com/wrench-project/wrench/tree/ec6b43561b95977002258c0fe37a4ecad8f1d33f/examples/basic-examples/io-pagecache}{ec6b43561b}.

            \begin{table}[b]
            \centering
            \begin{tabularx}{0.8\columnwidth}{c>{\centering\arraybackslash}X}
            \toprule
                Input size (GB)  & CPU time (second)\\
            \midrule
                3      & 4.4 \\
                20  & 28 \\
                50  & 75 \\
                75  & 110 \\
                100  & 155 \\
            \bottomrule
            \end{tabularx}
            \caption{Task CPU times used in the simulation.
%            \tristan{are the input files in GB or in GiB? If GiB, then change the abbreviation everywhere.}
%  Dzung: it's in GB
            \tristan{add nighres CPU times when available}}
            \label{table:cputime}
            \end{table}

@@

            Our first experiment runs a
            single instance of the synthetic application on a single
            cluster node, with different input file sizes (20~GB, 50~GB,
            75~GB, 100~GB), and with all I/Os directed to the same local
            disk. The goal is to validate the simulation model in absence
            of any concurrent disk access.

            Our second experiment runs concurrent instances of the
            application on a single node, all application instances
            operating on different files of size 3~GB stored in the same
            local disk. We vary the number of
            concurrent application instances from 1 to 32 since cluster
            nodes have 32 CPU cores.

            Our third experiment occurs in the same
            configuration as the previous one, albeit reading and writing
            on a 50-GiB NFS-mounted partition of a 450-GiB remote disk of
            another compute node. As is commonly configured in HPC
            environments to avoid data loss, there is no client write cache
            and the server cache is configured as writethrough instead of
            writeback. NFS client and server read caches are enabled. 
            Therefore, all the writes happen at disk bandwidth, but
            reads can benefit from cache hits.

            Our fourth and final experiment runs the Nighres workflow on
            data from participant 0027430 in the dataset of the Max Planck Institute for
            Human Cognitive and Brain Sciences
            available at
            \url{http://dx.doi.org/10.15387/fcp_indi.corr.mpg1}. The data
            amounts to 2.3~GB and was downloaded from
            \url{http://datasets.datalad.org/?dir=/corr/RawDataBIDS/MPG_1/sub-0027430}.
            The workflow was run on a single node of the cluster, using a
            single local disk.

@@

    \begin{figure*}
        \centering
        \begin{subfigure}{\linewidth}
            \centering
               \includegraphics[width=\linewidth]{result/single/figures/single_errors.pdf}
               \vspace*{-0.7cm}
               \caption{Absolute relative simulation errors}
               \vspace*{0.5cm}
               \label{fig:single_error}
            \end{subfigure}
        \begin{subfigure}{\linewidth}
            \centering
            %    Gray shades represent task phases (read, compute and write).
            %    Lines represent memory usage along pipeline execution time.}
               \includegraphics[width=\linewidth]{result/single/figures/single_memprof.pdf}
               \vspace*{-0.7cm}
               \caption{Memory profiles}
               \vspace*{0.5cm}
               \label{fig:single_memprof}
        \end{subfigure}
        \begin{subfigure}{\linewidth}
            \centering
               \includegraphics[width=1.05\linewidth]{result/single/figures/cached_files.pdf}
               \caption{Cache contents \emph{after} application I/O operations}
            %    \textcolor{red}{Update real results of 20GB}}
               \label{fig:single_cache}
        \end{subfigure}
        \caption{Single-threaded results}
        \end{figure*}

        \subsection{Single-threaded experiment}

@@

@@

        \begin{figure*}
            \begin{subfigure}{\linewidth}
                \centering
                \includegraphics[width=\linewidth]{result/multi/figures/multi_local.pdf}
            \end{subfigure}
            \caption{Concurrent results with 3~GB files (averages on 5 repetitions)}
            \label{fig:multi_local}
        \end{figure*}

        The content of the simulated memory cache is also highly consistent
        with reality (Figure~\ref{fig:single_cache}). With 20~GB files, the
        simulated cache content exactly matches reality, since all files fit in page cache. With 100~GB files, a slight
        discrepancy is observed after Write 2, which explains the
        simulation error previously mentioned in Read 3. In the real execution indeed,
        File 3 is entirely cached after Write 2, whereas in the simulated execution, 
        only a part of it is cached. 
        This is due to the fact that the Linux kernel tend to not evict 
        pages that belong to files being currently written (File 3 in this case), which we 
        could not easily reproduce in our model.
        \begin{figure*}
            \begin{subfigure}{\linewidth}
                \centering
                \includegraphics[width=\linewidth]{result/multi/figures/multi_nfs.pdf}
            \end{subfigure}
            \caption{NFS results with 3~GB files (averages on 5 repetitions)}
            \label{fig:multi_nfs}
            \end{figure*}

        \subsection{Concurrent applications}

@@

        \subsection{Remote storage (NFS)}

@@

        \subsection{Real application}

        \subsection{Overhead comparison}
        As is the case for \wrench, simulation time with \wrench-cache scales
        linearly with the number of concurrent applications
        (Figure~\ref{fig:multi_time}, p \textless $10^{-24}$). However, the page
        cache model substantially increases simulation time by
        application, as can be seen by comparing regression slopes in
        Figure~\ref{fig:multi_time}. Interestingly, \wrench-cache is faster with 
        NFS I/Os than with local I/Os, most likely due to the use of writethrough
        cache in NFS, which bypasses flushing operations.
        \begin{figure}
            \begin{subfigure}{\columnwidth}
                \centering
                \includegraphics[width=\linewidth]{result/multi/figures/simulation_time.pdf}
            \end{subfigure}
            \caption{Simulation time comparison. \wrench-cache scales
            linearly with the number of concurrent applications, albeit
            with a higher overhead than \wrench.}
            \label{fig:multi_time}
            \end{figure}

>>>>

        Our experiments compared real executions with our Python prototype,
        with the original \wrench simulator, and with our \wrench-cache
        extension. Executions included single-threaded and multi-threaded
        applications, accessing data on local and remote file systems. We
        used two applications: a synthetic one, created to evaluate the
        simulation model, and a real one, representative of neuroimaging
        data processing.

        Experiments were conducted on a dedicated cluster at
        Concordia University, with one login node, 9 compute nodes, and 4
        storage nodes connected with a network bandwidth of 25 Gbps. Each
        compute node had 2 $\times$ 16-core Intel(R) Xeon(R) Gold 6130 CPU
        @ 2.10GHz, 250~GiB of RAM, 6 $\times$ SSDs of 450~GiB each with the XFS
        file system, 378~GiB of tmpfs, and 126~GiB of devtmpfs file system.
        Nodes ran CentOS~8.1 with NFS version 4. We used the \texttt{atop}
        and \texttt{collectl} tools to monitor and collect memory status
        and disk throughput. We cleared the page
        cache before starting the application, to ensure comparable
        conditions among runs.

        \begin{table}[b]
            \centering
            \begin{tabularx}{0.8\columnwidth}{c>{\centering\arraybackslash}X}
            \toprule
                Input size (GB)  & CPU time (s)\\
            \midrule
                3      & 4.4 \\
                20  & 28 \\
                50  & 75 \\
                75  & 110 \\
                100  & 155 \\
            \bottomrule
            \end{tabularx}
            \caption{Synthetic application parameters}
            \label{table:cputime}
            \end{table}
        The synthetic application, implemented in C, consisted of three single-core,
        sequential tasks where each task read the file produced by the
        previous task, incremented every byte of this file to emulate real
        processing, and wrote the resulting data to disk. Files were
        numbered by ascending creation times: File 1 was the file read by Task 1, and file \texttt{i+1} was the file
        written by task \texttt{i}. The anonymous memory used by the application
        was released after each task, which we also simulated in the Python
        prototype and in WRENCH-cache. As our focus is on I/O rather than compute, we measured
        application task CPU times on a cluster node
        (Table~\ref{table:cputime}), and used these durations in our
        simulations. For the Python prototype, we injected CPU times
        directly in the simulation. For \wrench and \wrench-cache, we
        determined the corresponding number of flops on a 1~Gflops CPU
        and used these values in the simulation. The final simulated
        platform and application are available at
        commit \href{https://github.com/wrench-project/wrench/tree/ec6b43561b95977002258c0fe37a4ecad8f1d33f/examples/basic-examples/io-pagecache}{ec6b43561b}.

        We used the synthetic application in three experiments. In the
        first one (\textit{Exp 1}), we executed \emph{a single instance} of
        the application on a single cluster node, with different input file
        sizes (20~GB, 50~GB, 75~GB, 100~GB), and with all I/Os directed to
        the same local disk.
        In the second experiment (\textit{Exp 2}), we executed
        \emph{concurrent} instances of the application on a single node,
        all application instances operating on different files of size 3~GB
        stored in the same local disk. We varied the number of concurrent
        application instances from 1 to 32 since cluster nodes had 32 CPU
        cores.
        In the third experiment (\textit{Exp 3}), we used the same
        configuration as the previous one, albeit reading and writing
        on a 50-GiB NFS-mounted partition of a 450-GiB remote disk of
        another compute node. As is commonly configured in HPC
        environments to avoid data loss, there was no client write cache
        and the server cache was configured as writethrough instead of
        writeback. NFS client and server read caches were enabled. 
        Therefore, all the writes happened at disk bandwidth, but
        reads can benefit from cache hits.

        The real application was a workflow of the Nighres
        toolbox~\cite{huntenburg2018nighres}, implementing cortical
        reconstruction from brain images in four steps: skull stripping,
        tissue classification, region extraction, and cortical
        reconstruction. Each step read
        files produced by the previous step, and wrote files that were or were not read by the subsequent step. 
        More information on this
        application is available in the Nighres documentation at
        \url{https://nighres.readthedocs.io/en/latest/auto_examples/example_02_cortical_depth_estimation.html}.
        The workflow is implemented as a Python script that calls Java
        image-processing routines. We used Python 3.6, Java 8, and Nighres
        1.3.0. We patched the application to remove lazy data loading,
        which made CPU time difficult to separate from I/O time, and to
        monitor task CPU times, to inject them in the simulators. The
        patched code is available at
        \url{https://github.com/dohoangdzung/nighres}.
        \begin{table}
            \centering
            \begin{tabular}{lccc}
            \toprule
                \multicolumn{1}{c}{Workflow step}& Input size       & Output size      & CPU time\\
                                       & (MB)             & (MB)             & (s)\\
            \midrule
               Skull stripping         &  111             & 30               & 137 \\
               Tissue classification   &  22              & 175              & 624 \\
               Region extraction       &  175             & 121              & 81 \\
               Cortical reconstruction &  47              & 250              & 270\\
            \bottomrule
            \end{tabular} 
            \caption{Nighres application parameters}
            \label{table:nighres_stats}
            \end{table}

        We used the real application in the fourth and final experiment
        (\textit{Exp 4}). The workflow was run on a single node of the
        cluster, using a single local disk. We processed data from
        participant 0027430 in the dataset of the Max Planck Institute for
        Human Cognitive and Brain Sciences available at
        \url{http://dx.doi.org/10.15387/fcp_indi.corr.mpg1}, leading to the
        workflow parameters in Table~\ref{table:nighres_stats}.

        To parameterize the simulators, we benchmarked the
        memory, local disk, remote disk (NFS), and network bandwidths
        (Table~\ref{table:benchmark}). Since \simgrid, and thus \wrench, currently only supports
        symmetrical bandwidths, we use the mean of the read and write
        bandwidth values in our experiments.
            \begin{table}[b]
            \centering
            \begin{tabularx}{\columnwidth}{ll
            >{\centering\arraybackslash}X
            >{\centering\arraybackslash}X
            >{\centering\arraybackslash}X}
            \toprule
                \multicolumn{2}{c}{Bandwidths}  & Cluster (real) & Python prototype & \wrench simulator\\
            \midrule
            \multirow{2}{*}{Memory}      & read  & 6860 & 4812 & 4812\\
                                         & write & 2764 & 4812 & 4812\\
            \multirow{2}{*}{Local disk}  & read  & 510  & 465  & 465\\
                                         & write & 420  & 465  & 465\\
            \multirow{2}{*}{Remote disk} & read  & 515  & -    & 445\\
                                         & write & 375  & -    & 445\\
            \multicolumn{2}{l}{Network}  & 3000  & -    & 3000\\
            \bottomrule
            \end{tabularx}
            \caption{Bandwidth benchmarks (MBps) and simulator configurations.
            The bandwidths used in the simulations are the average of the read and write bandwidths
            measured in the cluster.
            Remote disk and network accesses are not simulated in the Python simulator.}
            \label{table:benchmark}
            \end{table}

@@

@@

    \begin{figure*}
        \centering
        \begin{subfigure}{\linewidth}
            \centering
               \includegraphics[width=\linewidth]{result/single/figures/single_errors.pdf}
               \vspace*{-0.7cm}
               \caption{Absolute relative simulation errors}
               \vspace*{0.5cm}
               \label{fig:single_error}
            \end{subfigure}
        \begin{subfigure}{\linewidth}
            \centering
            %    Gray shades represent task phases (read, compute and write).
            %    Lines represent memory usage along pipeline execution time.}
               \includegraphics[width=\linewidth]{result/single/figures/single_memprof.pdf}
               \vspace*{-0.7cm}
               \caption{Memory profiles}
               \vspace*{0.5cm}
               \label{fig:single_memprof}
        \end{subfigure}
        \begin{subfigure}{\linewidth}
            \centering
               \includegraphics[width=1.05\linewidth]{result/single/figures/cached_files.pdf}
               \caption{Cache contents \emph{after} application I/O operations}
            %    \textcolor{red}{Update real results of 20GB}}
               \label{fig:single_cache}
        \end{subfigure}
        \caption{Single-threaded results (\textit{Exp 1})}
        \end{figure*}

        \subsection{Single-threaded execution (Exp 1)}

@@

        \begin{figure*}
            \begin{subfigure}{\linewidth}
                \centering
                \includegraphics[width=\linewidth]{result/multi/figures/multi_local.pdf}
            \end{subfigure}
            \caption{Concurrent results with 3~GB files (\textit{Exp 2}). Averages on 5 repetitions.}
            \label{fig:multi_local}
        \end{figure*}

@@

        The content of the simulated memory cache is also highly consistent
        with reality (Figure~\ref{fig:single_cache}). With 20~GB files, the
        simulated cache content exactly matches reality, since all files fit in page cache. With 100~GB files, a slight
        discrepancy is observed after Write 2, which explains the
        simulation error previously mentioned in Read 3. In the real execution indeed,
        File 3 is entirely cached after Write 2, whereas in the simulated execution, 
        only a part of it is cached. 
        This is due to the fact that the Linux kernel tend to not evict 
        pages that belong to files being currently written (File 3 in this case), which we 
        could not easily reproduce in our model.
        \begin{figure*}
            \begin{subfigure}{\linewidth}
                \centering
                \includegraphics[width=\linewidth]{result/multi/figures/multi_nfs.pdf}
            \end{subfigure}
            \caption{NFS results with 3~GB files (\textit{Exp 3}). Averages on 5 repetitions.}
            \label{fig:multi_nfs}
            \end{figure*}

        \subsection{Concurrent applications (Exp 2)}

@@

        \subsection{Remote storage (Exp 3)}

@@

        \subsection{Real application (Exp 4)}
        As for the synthetic application, simulation errors are again
        substantially reduced by the WRENCH-cache simulator compared to
        WRENCH (Figure~\ref{fig:nighres}). On average, errors are reduced
        from 212~\% in WRENCH to 68~\% in WRENCH-cache. The error
        reduction is mostly due to more accurate read times, since write
        times remain generally identical between both implementations, presumably due to 
        the dominance of file compression time in write time.
        Similar to
        the synthetic application, the first read happens entirely from disk and is therefore 
        very accurately simulated by both WRENCH and WRENCH-cache.
        \begin{figure}
            \begin{subfigure}{\columnwidth}
                \centering
                \includegraphics[width=\linewidth]{result/nighres/figures/nighres_errors.pdf}
            \end{subfigure}
            \caption{Real application results (\textit{Exp 4})}
            \label{fig:nighres}
            \end{figure}

        \subsection{Simulation time}
        As is the case for \wrench, simulation time with \wrench-cache scales
        linearly with the number of concurrent applications
        (Figure~\ref{fig:multi_time}, p \textless $10^{-24}$). However, the page
        cache model substantially increases simulation time by
        application, as can be seen by comparing regression slopes in
        Figure~\ref{fig:multi_time}. Interestingly, \wrench-cache is faster with 
        NFS I/Os than with local I/Os, most likely due to the use of writethrough
        cache in NFS, which bypasses flushing operations.
        \begin{figure}
            \begin{subfigure}{\columnwidth}
                \centering
                \includegraphics[width=\linewidth]{result/multi/figures/simulation_time.pdf}
            \end{subfigure}
            \caption{Simulation time comparison. \wrench-cache scales
            linearly with the number of concurrent applications, albeit
            with a higher overhead than \wrench.}
            \label{fig:multi_time}
            \end{figure}
